---
title: "PS5: markdown, basic plots"
author: "Tianze Xu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r warning=FALSE, message=FALSE}
# load libraries
library(dplyr)
library(ggplot2)
```

## 1) Load and check data (5pt)

### You first task is to do a very simple data check:
### 1.1) (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating.

### 1.2) (2pt) Load data. How many rows/columns do we have?
```{r}
df <- read.csv('../data/gapminder.csv.bz2', sep = '\t')
nrow(df)
ncol(df)
```
There are 13055 rows and 25 columns.

### 1.3) (2pt) Print a small sample of data. Does it look OK?
```{r}
sample_n(df, 5)
```
Yes, the data looks okay.

## 2) Descriptive statistics (15pt)
### 2.1) (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name
```{r}
n_distinct(df$iso3)
n_distinct(df$iso2)
n_distinct(df$name)
```
There are 253 unique iso3 (3-letter country code), 249 unique iso2 (2-letter country code), and 250 unique country names.

### 2.2) If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more iso3 -codes. What is going on? Can you find it out?
#### 2.2.a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}
df %>% group_by(iso2) %>%
  summarise(name_count=n_distinct(name)) %>%
  arrange(desc(name_count))

df %>% filter(iso2=='') %>% select(iso3, iso2, name) %>% distinct()
```
There are 2 countries with missing iso2 values, these countries are CHANISL, GBM, KOS, NAM, and NLD_CURACAO.

#### 2.2.b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}
df %>% group_by(name) %>%
  summarise(iso3_count=n_distinct(iso3)) %>%
  arrange(desc(iso3_count))

df %>% filter(name=='') %>% select(iso3, iso2, name) %>% distinct()
```
There are 4 countries that have iso3-code, but without names. These countries are CHANISL, GBM, KOS, and NLD_CURACAO.

### 2.3) (2pt) What is the minimum and maximum year in these data?
```{r}
df %>% filter(!is.na(time)) %>%
  summarise(min = min(time),
            max = max(time))
```

## 3) CO2 emissions (30pt)
### Next, let’s analyze CO2 emissions.
### 3.1) (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?
```{r}
df %>% group_by(time) %>%
  summarise(co2_na_count = sum(is.na(co2) | is.na(co2_PC))) %>%
  arrange(desc(co2_na_count))
```
Year 2017, 2018, and 2019 have the most missing CO2 data.

### 3.2) (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
```{r}
# calculate co2 numbers by country
df_co2 <- df %>% filter(iso2 %in% c('US', 'CN', 'IN', 'SG', 'FR', 'CA') & !is.na(co2)) %>%
            group_by(name) %>%
            summarise(co2_total=sum(co2))

# bar plot using ggplot library
ggplot(data=df_co2, aes(x=name, y=co2_total)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 30))
```
From the above bar chart, we see US has the highest total CO2 emissions. China has the 2nd highest CO2 emissions. Singapore has the lowest CO2 emissions.

### 3.3) (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the same countries. What does this figure suggest?
```{r}
# calculate co2 numbers by country
df_co2_pc <- df %>% filter(iso2 %in% c('US', 'CN', 'IN', 'SG', 'FR', 'CA') & !is.na(co2_PC)) %>%
            group_by(name) %>%
            summarise(co2_pc_avg=mean(co2_PC))

# bar plot using ggplot library
ggplot(data=df_co2_pc, aes(x=name, y=co2_pc_avg)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 30))
```
The above figure suggests that US still has the highest CO2 emissions per capita, while India has the lowest CO2 emissions per capita.

### 3.4) (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
## Note: just compute averages over countries and ignore the fact that countries are of different size.
## Hint: Americas 2016 should be 4.80.
```{r}
# calculate co2 numbers by region
df_co2_pc_region <- df %>% filter(region!='' & !is.na(co2_PC)) %>%
                      group_by(region) %>%
                      summarise(co2_pc_avg=mean(co2_PC))

ggplot(data=df_co2_pc_region, aes(x=region, y=co2_pc_avg)) +
  geom_bar(stat="identity")
```
From the above plot, we see Europe is the continent that has the highest CO2 emissions per capita, while Africa has the lowest.

### 3.5) (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
## Hint: it should look something along these lines:
```{r}
df_co2_pc_region_year <- df %>% filter(region!='' & !is.na(co2_PC) & time %in% c(1960, 2016)) %>%
                      group_by(region, time) %>%
                      summarise(co2_pc_avg=mean(co2_PC))
df_co2_pc_region_year$Year <- as.factor(df_co2_pc_region_year$time)

ggplot(data=df_co2_pc_region_year, aes(x=region, y=co2_pc_avg, fill=Year)) +
  geom_bar(position = "dodge", stat="identity") +
  labs(x='Continent', y='Average CO2 per capita')
```

### 3.6) Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent)
```{r}
df %>% filter(time==2019 & !is.na(co2_PC))
```
Since all the co2_PC values in year 2019 are missing, we are not able to find the answer using this dataset.

## 4) GDP per capita (50pt)
### Let’s look at GDP per capita (GDP_PC).
### 4.1) (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.
### Comment what do you see there.
```{r}
df_1960 <- df %>% filter(time==1960) %>%
            group_by(name) %>%
            summarise(GDP_PC)
df_1960

#ggplot(mtcars, aes(x=wt, y=mpg)) +
 # geom_point(size=2, shape=23)
```


